# Data and Preliminary Analysis

## Data

The data, coming from Statistics Iceland, contain Icelandic population estimates and number of deaths for all ages $x = [0, 1, \dots, 104]$ and years $t = [1981, 1982, \dots, 2017]$. The ages were reduced into ten groups ($[0, 10)$, ..., $[80, 90)$ and $[90, 104]$) for ease of computation, leaving $10 \times 27 = 270$ age-year-observations of population size and deaths. Figure 1 shows the evolution of population size in the ten age-groups. We can see that the population composition has shifted noticably towards an older population.

```{r, fig.asp= 0.50, fig.cap = cap}
cap <- "Population sizes of year-age-groups, shown in log scale and overlaid on yearly totals for comparison."
d %>% 
  group_by(year) %>% 
  mutate(totalpop = sum(pop)) %>% 
  ungroup %>% 
  ggplot(aes(year, pop, fill = age)) +
  geom_area(aes(y = totalpop), alpha = 0.05, fill = "black", col = "grey90") +
  geom_area(col = "grey80") +
  scale_y_log10(breaks = c(1, 10, 100, 1000, 10000, 100000), 
                labels = comma,
                expand = expand_scale(mult = c(0, 0.1))) +
  scale_x_continuous(expand = expand_scale(mult = 0)) +
  scale_fill_brewer(type = "div", palette = "RdGy") +
  facet_wrap("age", ncol = 5) +
  guides(fill = "none") +
  labs(x = "Year", y = "Deaths") +
  background_grid(major = "none", minor = "none")
```

Figure 2 shows the evolution of number of deaths within each age-group over time. There has been a noticable decrease in deaths among the youngest part of the population, as well as an increase among the oldest part.

```{r, fig.asp= 0.50, fig.cap = cap}
cap <- "Number of deaths by year and age, shown in log scale and overlaid on yearly totals for comparison."
d %>% 
  group_by(year) %>% 
  mutate(totaldeaths = sum(deaths)) %>% 
  ungroup %>% 
  ggplot(aes(year, deaths, fill = age)) +
  geom_area(aes(y = totaldeaths), alpha = 0.05, fill = "black") +
  geom_area(col = "grey80") +
  scale_y_log10(breaks = c(1, 10, 100, 1000), 
                expand = expand_scale(mult = c(0, 0.1))) +
  scale_x_continuous(expand = expand_scale(mult = 0)) +
  scale_fill_brewer(type = "div", palette = "RdGy") +
  facet_wrap("age", ncol = 5) +
  guides(fill = "none", col = "none") +
  labs(x = "Year", y = "Deaths") +
  background_grid(major = "none", minor = "none")
```

This figure alone is not enough evidence for changes in mortality rates, but when population and deaths are used to calculate observed mortality rates, as is done in figure 3, there is a noticable rate of change in mortality rates, with each age-group seeming to have its own linear trend with added observational noise.

```{r, fig.cap = cap, fig.asp = 0.5}
cap <- "Time evolution of mortality rates by age-group, shown on log scale."
d %>% 
  mutate(label = ifelse(year == max(year), as.character(age), NA_character_),
         x = ifelse(is.na(label), NA, year + 2),
         y = case_when(label == "[30,40)" ~ rate - 0.00015,
                       label == "[40,50)" ~ rate + 0.00015,
                       label == "[0,10)" ~ rate - 0.00004,
                       label == "[20,30)" ~ rate + 0.00006,
                       TRUE ~ rate)) %>% 
  ggplot(aes(year, rate, col = age, group = age)) +
  geom_line(size = 0.6) +
  geom_text(aes(x = x, y = y, label = label)) +
  scale_y_log10(labels = percent, 
                breaks = c(0.1, 0.03, 0.01, 0.003, 0.001)) +
  scale_x_continuous(breaks = c(seq(1982, 2018, 4))) +
  scale_color_brewer(type = "div", palette = "RdGy") +
  guides(col = "none") +
  labs(x = "Year", y = "Mortality Rates")
```